Translation of B Implementations to the LLVM: Control and Modularity

### 2013-10-16 translation.tex
Many conceptual and implementation changes to implement support for imports. The prototype is not yet functional.

To encode the data space of a module, the translation produces a structure type,
named \llvm{\%M\$state\$}, such that the elements of the structure represent the
variables and the instances of imported modules. To satisfy our second initial
requirement (separate compilation), the instances cannot be part of the
structure itself, instead references to those instances (encoded as pointers)
are used. Let \llvm{\%M\$ref\$} denote the type pointer to \llvm{\%M\$state\$}.
To encode behavior of a module, each operation \B{op} is translated into the
definition of a function named \llvm{@M\$op}. The parameter list of the
functions thus produced contains one item for each input and output of the
corresponding operation. There is also one parameter of type \llvm{\%M\$ref\$},
to pass the address of the structure encoding instance associated with that
operation. Also the translation produces a special function \llvm{@M\$init\$}
responsible for executing the initialization of \B{M}. The parameters of this
function are the addresses of the instances found in the import tree of the
module (including the module itself).

Our approach consists in dividing the translation into three sections, and to
combine these sections differently in \emph{PROJ\/} and \emph{COMP\/} mode. For
a given module \B{M}, the sections are the following:
\begin{description}
\item[typedef:] Defines an LLVM data type \llvm{\%M\$state\$} corresponding to
  the data space of the module:
  \begin{tabbing}
    \llvm{\%M\$state\$ = type \{ \ListOf{\nt{type}} \}}
  \end{tabbing}
  In case the data space of a module is empty, this section is empty and the
  type is not defined.
\item[interface:] Defines an LLVM type \B{M\_ref}, pointer to \B{M\_data} (if
  defined):
  \begin{tabbing}
    \llvm{\%M\$ref\$ = type \%M\$state\$*}
  \end{tabbing}
  The interface also contains declarations of LLVM
  functions: one implementing the initialisation of the module, and one for each
  operation of the module. If the data space of the module is empty, no function
  definition corresponding to the initialisation is emitted:
  \begin{tabbing}
    foo \= \kill
    \llvm{declare void @M\$op(\%M\$ref\$, \ListOf{\nt{type}})} \\
    \llvm{declare void @M\$init\$(\%M\$ref\$, \ListOf{\nt{type}})}
  \end{tabbing}
\item[implementation:] Defines the functions declared in the interface:
  \begin{tabbing}
    foo \= \kill
    \llvm{define void @M\$op(\%M\$state\$* \%self\$, \ListOf{\nt{param}}) \{} \\
    \> \llvm{\ListOf{\nt{block}}} \\
    \> \llvm{exit: ret void} \\
    \llvm{\}}
  \end{tabbing}
  This section is produced only for the developed modules.
\end{description}
Based on the definition of these sections, the \emph{COMP\/} translation
assembles them as follows:
\begin{center}
  \begin{tabbing}
    foo \= foo \= \kill
    \textit{for each imported module} \B{Q} \textit{generate} \\
    \> \llvm{\%Q\$state\$ = type opaque} \quad \textit{(declares type for} \B{Q} \textit{state space)} \\
    \> \textit{include the interface section of }\B{Q} \\
    \textit{include the typedef section of} \B{M} \\
    \textit{include the implementation section of} \B{M}
  \end{tabbing}
\end{center}
For a module \B{M}, the \emph{PROJ\/} translation is organized as follows:
\begin{center}
  \begin{tabbing}
    foo \= foo \= \kill
    \textit{for each transitively imported component} \B{Q} \\
    \> \textit{include the typedef section of \B{Q}} \\
    \textit{for each instance} \B{Q} \textit{imported transitively through} \B{path} \\
    \> \textit{declare a variable of type} \llvm{\%Q\$state\$}: \\
    \> \llvm{@Q$\lbrack$path$\rbrack$ = common global \%M\$state\$ zeroinitializer} \\

>>>>

To encode the data space of a module, the translation produces a structure type,
named \llvm{\%M\$state\$}, such that the elements of the structure represent the
variables and the instances of imported modules. To satisfy our second initial
requirement (separate compilation), the instances cannot be part of the
structure itself, instead references to those instances (encoded as pointers)
are used. Let \llvm{\%M\$ref\$} denote the type pointer to \llvm{\%M\$state\$}.
To encode behavior of a module, each operation \B{op} is translated into the
definition of a function named \llvm{@M\$op}. The parameter list of the
functions thus produced contains one item for each input and output of the
corresponding operation. There is also one parameter of type \llvm{\%M\$ref\$},
to pass the address of the instance encoding structure associated with that
operation. Also the translation produces a special function \llvm{@M\$init\$}
responsible for executing the initialization of \B{M}. The parameters of this
function are the addresses of the instances found in the import tree of the
module (including the module itself).

Our approach consists in dividing the translation into three sections, and to
combine these sections differently in \emph{PROJ\/} and \emph{COMP\/} mode. For
a given module \B{M}, the sections are the following:
\begin{enumerate}
\item typedef: Defines an LLVM data type \llvm{\%M\$state\$} corresponding to
  the data space of the module:
  \begin{tabbing}
    \llvm{\%M\$state\$ = type \{ \ListOf{\nt{type}} \}}
  \end{tabbing}
  In case the data space of a module is empty, this section is empty and the
  type is not defined.
\item interface: Defines an LLVM type \B{M\_ref}, pointer to \B{M\_data} (if
  defined):
  \begin{tabbing}
    \llvm{\%M\$ref\$ = type \%M\$state\$*}
  \end{tabbing}
  The interface also contains declarations of one LLVM functions for each
  operation of the module. 
  \begin{tabbing}
    \llvm{declare void @M\$op(\%M\$ref\$, \ListOf{\nt{type}})}
  \end{tabbing}
  If the data space of the module is not empty, a function definition
  corresponding to the initialisation is also emitted:
  \begin{tabbing}
    \llvm{declare void @M\$init\$(\%M\$ref\$, \ListOf{\nt{type}})}
  \end{tabbing}
\item implementation: Defines the functions declared in the interface:
  \begin{tabbing}
    foo \= \kill
    \llvm{define void @M\$op(\%M\$state\$* \%self\$, \ListOf{\nt{param}}) \{} \\
    \> \llvm{\ListOf{\nt{block}}} \\
    \> \llvm{exit: ret void} \\
    \llvm{\}}
  \end{tabbing}
  This section is produced only for the developed modules.
\end{description}
Based on the definition of these sections, the \emph{COMP\/} translation
assembles them as follows:
\begin{center}
  \begin{tabbing}
    foo \= foo \= \kill
    \textit{for each imported module} \B{Q} \textit{generate} \\
    \> \llvm{\%Q\$state\$ = type opaque} \quad \textit{(declares type for} \B{Q} \textit{state space)} \\
    \> \textit{include the interface section of }\B{Q} \\
    \textit{include the typedef section of} \B{M} \\
    \textit{include the implementation section of} \B{M}
  \end{tabbing}
\end{center}
For a module \B{M}, the \emph{PROJ\/} translation is organized as follows:
\begin{center}
  \begin{tabbing}
    foo \= foo \= \kill
    \textit{for each transitively imported component} \B{Q} \\
    \> \textit{include the typedef section of \B{Q}} \\
    \textit{for each instance} \B{Q} \textit{imported transitively through} \B{path} \\
    \> \textit{declare a variable of type} \llvm{\%Q\$state\$}: \\
    \> \llvm{@Q$\lbrack$path$\rbrack$ = common global \%M\$state\$ zeroinitializer} \\
